Exploiting hydraulic geometry to remotely sense gas exchange velocity via the SWOT satellite

### 2022-06-08 docs/intial_drafts/manuscript_v6.Rmd
first round of review, setting up for rerun on HPC

---
title: "Gas exchange in large rivers influenced by multiple turbulence scales: implications for remotely sensing gas exchange via the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Peter A Raymond:
      institute: Yale
  - Christopher J Zappa:
      institute: Lamont
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts-Amherst
  - Yale: School of the Environment, Yale University
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

## Key Points
- Multiple scales of turbulence influence gas exchange in the large rivers that the SWOT satellite will measure
- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT data and in situ $CO_2$ data
- BIKER is also robust to measurement errors implicit in SWOT river measurements

## Keywords
gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry, ungauged basin

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gases to the atmosphere via evasion. Despite much progress, there is still significant uncertainty in the temporal dynamics of gas exchange velocity (the rate of this evasion), with most models relying on temporally-coarse estimates of this velocity. We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of river hydrology research at the global scale, will provide similarly transformative insights to fluvial gas exchange modeling. Here, we present work inferring gas exchange velocity from SWOT observations. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model that is nearly entirely remotely sensible. We then couple this model with established Bayesian techniques to develop BIKER, or the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 763 discrete measurements of river gas exchange, yielding 1) a theoretical model that explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of variation in gas exchange in SWOT-like rivers and 2) an algorithm that predicts this theoretical model solely from SWOT observations with a by-river median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%. BIKER is most robust at reproducing the temporal dynamics of gas exchange (by-river median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), rather than estimating the absolute rates of exchange. Finally, it is robust to measurement errors implicit in the SWOT data. With SWOT set to launch in 2022, we suggest that BIKER will prove a fundamental asset for exploring global-scale, spatiotemporal dynamics of fluvial gas exchange.

## 1 Introduction
Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Since the publication of @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] (*Liu etal in review- add number is paper is accepted in time*), with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite its incredibly small percentage of the global land surface [0.`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`%- @raymondGlobalCarbonDioxide2013], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011]. It is still relatively poorly constrained (as evident by these studies' reported uncertainties) but is clearly a critical component of the global carbon cycle. Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used throughout this study.

$$FCO_2=k(CO_{2_{water}}-CO_{2_{air}}) \mathbf{(1)}$$

The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020]. Even beyond $CO_2$ emissions, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g. @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008]. Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009 for a review] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020 for a review].

Given that fluvial *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and laboratory experiments have converged on the 'small-eddy model' as first proposed by @lamontEddyCellModel1970 which scales *k* via the smallest-scale turbulent eddies [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @katulStructureFunctionModel2018]. This model is provided as equation 2, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number. To compare across gases and water temperatures, *k* is often reported as $k_{600}$, where *k* is normalized to a Schmidt number of 600 [reflecting a $k_{CO_2}$ at 20 degrees Celsius- @hallGasExchangeStreams2020]. Note that directly measuring $\epsilon$ is non-trivial; therefore, a frequently used model for $\epsilon$ assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993].

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(2)}$$

While this model works reasonably well in non-fluvial environments where turbulence production is mostly limited to bottom shear [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainty in how it applies to fluvial systems. In particular, there is uncertainty as to which models are most appropriate at different locations along the stream-to-river continuum given that river hydraulics change predictably along a river's longitudinal profile [@wangPhysicallyBasedScaling2021]. In small and steep headwaters, 'bubble-mediated gas exchange' in whitewater might lead to substantially elevated *k* where bedload is large and general channel roughness is extremely high [@mauriceInfluenceFlowBed2017; @ulsethDistinctAirWater2019]. At the other end of the stream-to-river continuum (the largest rivers draining the continents), relatively little attention has been paid to gas exchange. The handful of existing field studies of large-river *k* suggest that *k* begins to be influenced by wind shear once the water surface is less protected, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021]. Other workers have noted that laboratory and field observations suggest small-roughness open channel flows (generally big rivers) do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. @moogAirWaterGasTransfer1999 proposed a conceptual 'chainsaw model', where large, depth-scale eddies transport small eddies to the interface, but the small eddies are responsible for the actual gas evasion that occurs at this interface. Finally, another approach proposed regardless of one's location along the stream-to-river continuum is to model *k* via 'form-drag shear' (i.e. total stream power per unit mass water). This normalized stream power captures the bulk frictional resistance via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that this model can reasonably predict *k* in rivers [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021], however it requires estimates of the mean flow velocity and therefore can introduce additional uncertainties.

Taken in aggregate, our theoretical understanding of fluvial gas exchange is likely limited by a dearth of field-measured fluvial *k*, particularly in large rivers where direct measurements are difficult. @wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 reasonably sized rivers (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to implicitly solve for *k*. They found that equation 2, coupled with $\epsilon$ modeled via the log-law-of-the-wall, is valid in their simulated rivers and that *k*~streamflow relationships break down in large rivers. However, their simulated rivers were limited by a lack of direct hydraulic measurements and had to rely on models to estimate river depth and velocity, introducing additional uncertainty. Further, they did not assess model fit in very steep systems [@ulsethDistinctAirWater2019] and stopped short of parsing out hydraulic explanations for their results.

All of these mechanistic uncertainties are then propagated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] via equation 1 coupled with hydraulic geometry (HG). HG is the scaling relationships between streamflow and river channel hydraulics [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] that allow biogeochemists to predict river depth and velocity from streamflow models and/or streamgauges . It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker. Further, these approaches rely on either in situ discharge records or modeled streamflow which further contributes uncertainty. This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

A potential alternative to this upscaling approach is to directly estimate a river's hydraulic properties (and thus *k*) from remote sensing (RS) data. Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge [RSQ- @gleasonRemoteSensingRiver2020]. This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on hydraulic generalizability in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2020]. Many, but not all, of these ungauged approaches are developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle and will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].

Once SWOT launches, there will be a massive trove of high-temporal-resolution hydraulics measurements of the world's largest river surfaces, as well as established techniques to infer streamflow from those measurements. Given the uncertainties and limited field measurements of river *k* as described above, the SWOT measurements will hopefully also prove transformative in elucidating the spatiotemporal dynamics of gas exchange. However, in order to make sense of SWOT data for these purposes, a method must first be developed that can ingest SWOT data and return reasonable estimates of the gas exchange velocity solely from those data. In this context, we borrow tools from fluvial geomorphology and the SWOT discharge algorithms to answer the following questions: 1) can we leverage the unique geomorphology of the large rivers that SWOT will observe to develop a physically-based gas evasion model unique to these rivers? And 2) can we then exploit the findings from this analysis to develop a novel methodology that predicts this physically-based $k_{600}$ model (and its explicit uncertainty) solely using SWOT observations?

We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference of the $k_{600}$ Evasion Rate' and it requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched). We also quantify BIKER's sensitivity to the expected SWOT measurement errors on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$ from streamflow.

## 2 Paper overview
Following the description of the data used (section 3), this paper is split into two distinct parts: gas-exchange model development and validation (section 4) and BIKER algorithm development and validation (section 5). Each section is further split into both methods and results sections, with section 5 fundamentally dependent on section 4. We conclude with a Discussion regarding both sections 4 and 5 (section 6). We also provide a flowchart detailing the entire study as Figure 1.

@@

## 3 Data
Numerous datasets were used in this study. Please see Figure 2 for a map of the approximate locations for the data used in this study.

@@

Gas exchange theory (section 4) is explored using multiple datasets where $k_{600}$ was measured directly and/or collected from the literature. In total, there are 763 measurements of stream hydraulics and $k_{600}$. This is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$. 701 of these measurements were previously collected from other studies by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining measurements were added by us but measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements.

To quantify our geomorphic assumptions for SWOT rivers (section 4.1), we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. These were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States. Ultimately, this dataset was filtered down to `r sum(percs$n)` discrete measurements for this study (see Text S1 for our filtering protocol).

BIKER validation (section 4) was performed on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers. Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on "SWOT-like" data. These simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER. Here, we use `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016. These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission. The approximate locations of these rivers are plotted in Figure 1, and please consult both of those papers for all of the hydraulic model specifications. Ultimately, the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh. We omit three models from @durandIntercomparisonRemoteSensing2016 because they lack enough hydraulic information to calculate the mean flow depth, which is necessary for algorithm validation (section 5.1.2). These are two models for different sections of the Saint Lawrence River in Canada and one for the Tanana River in Alaska, United States.

To assess the influence of SWOT's implicit measurement error on BIKER performance, we use the error model developed by @durandHowWillRadar2020 and implemented on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])`/`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` of the rivers by @frassonExploringFactorsControlling2021. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020]. Width errors due to poor water classification are ignored as they were in @frassonExploringFactorsControlling2021. @frassonExploringFactorsControlling2021 tested a suite of measurement error scenarios, however here we test only their 'full uncertainty scenario' that most closely represents the measurement errors likely to be implicit in SWOT data.

For the $CO_2$ evasion and carbon efflux calculations (section 5.1.3), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S1). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.

## 4 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
Given that *k* can only be directly measured via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020], SWOT data will not provide the necessary parameters to directly measure *k*. Further, these techniques are entirely infeasible at the global scale where our goal with BIKER is primarily to better understand spatiotemporal patterns of *k* rather than obtain absolute measurements of it. Therefore, we need a simple yet physically-based model for *k* to directly relate channel hydraulics to gas exchange in the large rivers that SWOT will measure. We hypothesize that this can be done by leveraging the unique geomorphology of the very wide rivers that SWOT will observe. In this section, we develop and test process-based models for *k*, selecting one that is then implemented within BIKER (section 5).

>>>>

---
title: "Exploiting hydraulic geometry to remotely sense gas exchange velocity via the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Peter A Raymond:
      institute: Yale
  - Christopher J Zappa:
      institute: Lamont
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts-Amherst
  - Yale: School of the Environment, Yale University
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

## Key Points
- BIKER algorithm predicts gas exchange velocity and $CO_2$ fluxes from simulated SWOT data
- BIKER is robust to measurement errors implicit in SWOT river measurements
- BIKER will allow for unprecedented study of gas exchange spatiotemporal dynamics upon SWOT's launch

## Keywords
gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via evasion. Despite much progress, there is still significant uncertainty in the temporal dynamics of gas exchange velocity (the rate of evasion) and thus carbon emissions to the atmosphere. We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring gas exchange velocity from future SWOT observations without an in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of variation in gas exchange. We then couple this model with established Bayesian techniques to develop BIKER, or the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 763 discrete measurements of river gas exchange, yielding an algorithm that predicts this theoretical model solely from SWOT observations with a by-river median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%. Like many fluvial Bayesian inference algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), rather than estimating the absolute rates of exchange. Finally, we show BIKER robust to measurement errors implicit in the SWOT data. With SWOT set to launch in 2022, we suggest that BIKER can be deployed to map the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

## 1 Introduction
Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to their long understood role of exporting carbon to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] (*Liu etal in review- add number is paper is accepted in time*), with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite its incredibly small percentage of the global land surface [0.`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`%- @raymondGlobalCarbonDioxide2013], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011]. River $CO_2$ evasion is increasingly better constrained (as evident by these studies' reported uncertainties) and is clearly a critical component of the global carbon cycle. Equation 1 represents this flux given $\delta CO_2$ (the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$) and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used throughout this study.

$$FCO_2=k* \delta CO_2 \mathbf{(1)}$$
The structure of equation 1 necessitates that calculations of this flux are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly measured via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. When trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this limitation necessitates the use of predictive models for fluvial *k* that are based on easily estimated river hydraulic properties. In that vein, there have been over 20 empirical models developed for this and similar purposes, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@wangPhysicallyBasedScaling2021]. Note that these models usually predict $k_{600}$, or *k* normalized by a Schmidt number of 600, to make use of additional measurements taken at different water temperatures and with different gases. For example, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius.

While this workflow enables estimating the global $FCO_2$ flux ($FCO_2_{upscaled}$) across millions of rivers, it also changes the base parameters that ultimately control that estimate. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$ [of which an extensive literature has explored- e.g. @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000], but also the accuracy and resolution of our river hydraulics estimates.

$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
For example, the @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 $FCO_2$ global upscaling studies all relied on $k_{CO_2}$ values calculated using mean annual hydraulic conditions, while @borgesGloballySignificantGreenhousegas2015 used a combination of mean annual values and a constant $k_{CO_2}$ in space and time to upscale over Africa. In all of these key studies, the temporal dynamics of the gas exchange velocity (and thus $FCO_2$) were ignored because of data limitations. As workers move towards river-reach-explicit modeling of gas exchange at the global scale with more and more elaborate models for $CO_{water}$ [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; *Liu etal in review; Saccardi & Winnick et al in review*], they will also have to contend with the lack of direct measurements of global river hydraulics.

This lack of direct measurements also inhibits our process-level understandings of the spatiotemporal dynamics of $k_{600}$. A few field-scale studies have directly investigated these dynamics but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012]. For example, @wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to a mean of only 8 measurements per river in a single watershed. They also often focus on small rivers. More recently, @wangPhysicallyBasedScaling2021 attempted to address this problem by simulating *k* in 35 reasonably sized rivers (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to implicitly solve for *k*. They found that process-based models for *k* generally work well in these simulated rivers and that *k*~streamflow relationships change with river size, though they too were limited by a lack of direct hydraulic measurements and had to rely on scaling equations to estimate river depth and velocity.

In theory, the above problems could be alleviated and a global analysis of *k* spatiotemporal trends would be possible if direct hydraulics measurements were available at the global scale at a sufficient temporal resolution. Conveniently, near-daily hydraulics measurements will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulics measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and fluvial biogeochemsitry should be no different. We suggest that SWOT data will prove transformative to global-scale biogeochemistry if techniques are developed that can ingest SWOT data and infer the parameters of interest. Here, we focus specifically on $k_{600}$, borrowing tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

  - Can we leverage the unique geomorphology of the large rivers that SWOT will observe to develop a physically-based gas evasion model unique to these rivers?

  - Can we then exploit the findings from this analysis to develop a novel methodology that infers this physically-based $k_{600}$ model (and its uncertainty) solely using SWOT observations?

We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference of the $k_{600}$ Evasion Rate' and it requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched). We also quantify BIKER's sensitivity to the expected SWOT measurement errors on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of those rivers: SWOT data is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors as a function of the cross-track distance [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$ from streamflow.

Following the description of the data used (section 2), this paper is split into two distinct parts: gas-exchange theory and model development (section 3) and BIKER algorithm development and validation (section 4). Section 4 is fundamentally dependent on the results presented in section 3. We conclude with a Discussion regarding both sections 3 and 4 (section 5). We also provide a flowchart detailing the entire study as Figure 1.

@@

## 2 Data
Numerous datasets were used in this study. Please see Figure 2 for a map of the approximate locations for the data used in this study.

@@

Gas exchange theory (section 3) is explored using multiple datasets where $k_{600}$ was measured directly and/or collected from the literature. In total, there are 763 measurements of stream hydraulics and $k_{600}$. This is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$. 701 of these measurements were previously collected from other studies by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining measurements were added by us but measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements.

To quantify our geomorphic assumptions for SWOT rivers (section 3.1), we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. These were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States. Ultimately, this dataset was filtered down to `r sum(percs$n)` discrete measurements for this study (see Text S1 for our filtering protocol).

BIKER validation (section 4) was performed on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers. Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on "SWOT-like" data. These simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER. Here, we use `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016. These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission. The approximate locations of these rivers are plotted in Figure 1, and please consult both of those papers for all of the hydraulic model specifications. Ultimately, the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh. We omit three models from @durandIntercomparisonRemoteSensing2016 because they lack enough hydraulic information to calculate the mean flow depth, which is necessary for algorithm validation (section 4.1.2). These are two models for different sections of the Saint Lawrence River in Canada and one for the Tanana River in Alaska, United States.

To assess the influence of SWOT's implicit measurement error on BIKER performance, we use the error models developed by @durandHowWillRadar2020 and @frassonExploringFactorsControlling2021, which the latter implemented on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])`/`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` of the rivers. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error and river width error. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020]. Width errors due to poor water classification are ignored but width errors due to the cross-track distance are included [@frassonExploringFactorsControlling2021]. The cross-track distance, reflective of the amount of the river cross-section that falls within SWOT's swath, controls the size of the pixels **help, I on't understand cross-track distance**. @frassonExploringFactorsControlling2021 tested a suite of measurement error scenarios, however here we test only their 'full uncertainty scenario' that most closely represents the measurement errors that will be implicit in SWOT data.

For the $CO_2$ evasion and carbon efflux calculations (section 4.1.3), we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S1). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.

## 3 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in large rivers, and then obtain a suite of models we can test. First, we outline these established models (section 3.1) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 3.2). Following this model development, all models are empirically tested on the dataset described in Section 2 after filtering for measurements made in hydraulically-wide channels (section 3.3). This was defined here as those whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to 166 direct measurements of hydraulically-wide $k_{600}$ to test with.
